Modeling Second-Language Learning from a Psychological Perspective

### 2018-03-29 doc/naaclhlt2018-latex/paper.tex
ok

Educational software that aims to teach people new skills, languages, and
academic subjects have become increasingly popular.  The wide-spread
deployment of these tools has created interesting opportunities to study
the process of language acquisition in extremely large samples of learners in naturalistic
situations. The Duolingo shared task on Second Lanugage Acquisition Modeling (SLAM)
was a competitive modeling challenge run in early 2018~\cite{slam18}.
The challenge, organized by Duolingo\footnote{http://duolingo.com}, a popular second
language learning app, was
to use log data from thousands of users completing millions of exercises to 
predict patterns of future translation mistakes in held-out data.  The data was
divided into three sets covering Spanish speakers learning English ({\tt en\_es}),
English speakers learning Spanish ({\tt es\_en}), and English speakers learning
French ({\tt fr\_en}).
This paper reports the approach used by our team which ended in third place 
for the {\tt en\_es} data set, second place for {\tt es\_en}, and third place
for {\tt fr\_en}.

@@

Despite these promising opportunities, it remains unclear how much of current psychological
theory might be important for tasks such as the Duolingo SLAM challenge.
In the field of education data mining, researchers trying to build
predictive models of student learning have typically relied on traditional, and interpretable, models and approaches
that are rooted in cognitive science~\cite[e.g.,][]{Atkinson:1972rm,Atkinson:1972rz,Corbett1995bkt,Pavlik:2008rm}.
However, a recent paper found that state-of-the-art results could be achieved using deep neural networks with
little or no cognitive theory built in~\cite[so called ``deep knowledge tracing",][]{Piech:2015aa} 
In response, a number of papers such as Khajah, Lindsey, \& Mozer (2016) compared deep knowledge tracing (DKT) to more
standard ``Bayesian knowledge tracing" (BKT) models and showed
that it was possible to equate the performance of the the BKT model by additional 
features and parameters that represent core aspects of the psychology of learning
and memory such as forgetting and individual abilities~\cite{Khajah2016hdnt}. An ongoing debate remains in this community 
whether using flexible models with lots of data can improve over more heavily structured, theory-based models \cite{Tang:2016:DNN, xiong2016going, zhang2017incorporating}.

For our approach to the SLAM competition, we decided to use a generic and fairly flexible model structure that we provided with hand-coded, psychologically inspired features.
We therefore positioned our entry to SLAM somewhat in between the approaches mentioned above.
Specifically, we used gradient boosting decision trees~\cite[GBDT][]{ke2017lightgbm} for the model structure, which is a powerful classification algorithm that is known to perform well across various kinds of data sets.  Like
deep learning, GBDT can extract complex interactions among features,
but it has some advantages including faster training and easier integration of
diverse inputs.

We then created a number of new psychologically-grounded features for the SLAM
dataset covering aspect such as user perseverance, learning processes, 
contextual factors, and cognate similarity.  After finding a model that provided the best
held-out performance on the test data set, we conducted a number of ``lesioning" studies where
we selectively removed features from the model and re-estimated the parameters in order
to assess the contribution of particular types of features.  We begin by describing our 
overall modeling approach, and then discuss some of the lessons learned from our analysis.

@@

First, we encoded the number of times the
current exercise's exact sentence had been seen before by the user. This is
informed by psychological research showing memory and perceptual processing
improvement for repeated contexts or chunks~\cite[e.g.,][]{Chun:1999gt}.

@@

To tune model hyper-parameters and evaluate the usefulness of features, we first
trained the models on the {\tt train} data set and evaluated them on the {\tt dev}
data set. \hl{Anselm: if space permits, mention what the train, dev, and test data sets look like? e.g.:  80\% of each user's exercises are in train, 10\% in each of dev and test.} Once the model structure was finalized, we trained on the combined
{\tt train} and {\tt dev} data and produced predictions for the {\tt test} data. The
LightGBM hyperparameters used for each model are listed in Table~\ref{lightgbm-params}.

@@

\hl{Anselm: mention AUROC of baseline model for comparison? shall the reader be impressed by our performance. perhaps also mention the winner's score (the deep learning team) for commparison, to show that our score was not that much worse? `we were 0.01 behind the deep learning model' sounds cooler to me than `we were third'}
The AUROC of our final predictions was $.8585$ on {\tt en\_es}, $.8350$ on {\tt
  es\_en}, and $.8540$ on {\tt fr\_en}. We did not attempt to optimize the model's
F1 score, but the F1 score could likely be improved (at the cost of
increased log loss) by finding the
rescaling of the {\tt dev} predicted probabilities that maximized the F1 score at
the 0.5 threshold, and applying this rescaling to the {\tt test} predicted probabilities.

@@

Though not plotted above, we also ran a model lesioning exercise-level features
including client, session type, format, and exercise duration. This model
achieved an AUROC of $.787$, far lower than any other lesion. This points to the
importance of how a question is asked for user performance, reflecting insights
from psychology such as the difference between recall and recognition memory
\cite{yonelinas2002nature}. 
\hl{Anselm: I don't understand the last sentence here.}

@@

Finally, if the task had required broader generalization, psychologically inspired 
features might have performed more competitively. In the SLAM
task, there is a large amount of labeled training data for every user and for
most words. This allows simple ID-based features to work because the past
history of a user will likely influence their future performance. However, with
 ID-based features there is no way to generalize to newly-encountered users or 
 words, which have an ID that was not in the training set. The learned 
ID-based knowledge is useless here because there is no way to generalize from one 
unique ID to another. Theory-driven features, in
contrast, can often generalize to new settings because they capture aspects that 
are shared across (subsets of) users, words, or situations of the learning task.  For example, if we were asked to generalize to
a completely new language such as German, many parts of our model would falter
but word frequency, age of acquisition, and Levenshtein distance to
first-language translation would still likely prove to be features which
have high predictive utility.
\hl{I like that sentence! Bam take that, deep learning}

In sum, we believe that the Duolingo SLAM dataset and challenge provide
interesting opportunities for cognitive science and psychology. 
Large-scale, predictive challenges like this one
might be used to identify features or variables that are important for learning.
Then, complementary laboratory-scale studies can be conducted which
establish the causal status of such features through controlled experimentation.
Conversely, insights from controlled experiments can be used to generate
new features that aid predictive models on naturalistic datasets. This type of
two-way interaction could lead to long-run improvements in both scientific
explanation and real-world prediction.

>>>>

Educational software that aims to teach people new skills, languages, and
academic subjects have become increasingly popular.  The wide-spread
deployment of these tools has created interesting opportunities to study
the process of learning in large samples. The Duolingo shared task on Second Lanugage Acquisition Modeling (SLAM)
was a competitive modeling challenge run in early 2018~\cite{slam18}.
The challenge, organized by Duolingo\footnote{http://duolingo.com}, a popular second
language learning app, was
to use log data from thousands of users completing millions of exercises to 
predict patterns of future translation mistakes in held-out data.  The data was
divided into three sets covering Spanish speakers learning English ({\tt en\_es}),
English speakers learning Spanish ({\tt es\_en}), and English speakers learning
French ({\tt fr\_en}).
This paper reports the approach used by our team which ended in third place 
for the {\tt en\_es} data set, second place for {\tt es\_en}, and third place
for {\tt fr\_en}.

@@

Despite these promising opportunities, it remains unclear how much of current psychological
theory might be important for tasks such as the Duolingo SLAM challenge.
In the field of education data mining, researchers trying to build
predictive models of student learning have typically relied on traditional, and interpretable, models and approaches
that are rooted in cognitive science~\cite[e.g.,][]{Atkinson:1972rm,Atkinson:1972rz,Corbett1995bkt,Pavlik:2008rm}.
However, a recent paper found that state-of-the-art results could be achieved using deep neural networks with
little or no cognitive theory built in~\cite[so called ``deep knowledge tracing",][]{Piech:2015aa}.
Khajah, Lindsey, \& Mozer (2016) compared deep knowledge tracing (DKT) to more
standard ``Bayesian knowledge tracing" (BKT) models and showed
that it was possible to equate the performance of the the BKT model by additional 
features and parameters that represent core aspects of the psychology of learning
and memory such as forgetting and individual abilities~\cite{Khajah2016hdnt}. An ongoing debate remains in this community 
whether using flexible models with lots of data can improve over more heavily structured, theory-based models \cite{Tang:2016:DNN, xiong2016going, zhang2017incorporating}.

For our approach to the SLAM competition, we decided to use a generic and fairly flexible model structure that we provided with hand-coded, psychologically inspired features.
We therefore positioned our entry to SLAM somewhat in between the approaches mentioned above.
Specifically, we used gradient boosting decision trees~\cite[GBDT,][]{ke2017lightgbm} for the model structure, which is a powerful classification algorithm that is known to perform well across various kinds of data sets.  Like
deep learning, GBDT can extract complex interactions among features,
but it has some advantages including faster training and easier integration of
diverse inputs.

We then created a number of new psychologically-grounded features for the SLAM
dataset covering aspects such as user perseverance, learning processes, 
contextual factors, and cognate similarity.  After finding a model that provided the best
held-out performance on the test data set, we conducted a number of ``lesioning" studies where
we selectively removed features from the model and re-estimated the parameters in order
to assess the contribution of particular types of features.  We begin by describing our 
overall modeling approach, and then discuss some of the lessons learned from our analysis.

@@

First, we encoded the number of times the
current exercise's exact sentence had been seen before by the user. This is
informed by psychological research showing memory and perceptual processing
improvements for repeated contexts or ``chunks"~\cite[e.g.,][]{Chun:1999gt}.

@@

To tune model hyper-parameters and evaluate the usefulness of features, we first
trained the models on the {\tt train} data set and evaluated them on the {\tt dev}
data set. Details of the training sets and the actual files are provided on the
Harvard Dataverse\footnote{https://dataverse.harvard.edu/dataset.xhtml?persistentId=doi:10.7910/DVN/8SWHNO}.
Once the model structure was finalized, we trained on the combined
{\tt train} and {\tt dev} data and produced predictions for the {\tt test} data. The
LightGBM hyperparameters used for each model are listed in Table~\ref{lightgbm-params}.

@@

The AUROC of our final predictions was $.8585$ on {\tt en\_es}, $.8350$ on {\tt
  es\_en}, and $.8540$ on {\tt fr\_en}. For reference this placed us within .01 of the winning
  entry for each problem ($.8613$ on {\tt en\_es}, $.8383$ on {\tt
  es\_en}, and $.8570$ on {\tt fr\_en}).  Also note that the Duolingo-provided baseline
  model (L2-regularized regression trained with stochastic gradient descent weighted
  by frequency) obtains $.7737$ on {\tt en\_es}, $.7456$ on {\tt
  es\_en}, and $.7707$ on {\tt fr\_en}.
  We did not attempt to optimize the model's
F1 score, but the F1 score could likely be improved (at the cost of
increased log loss) by finding the
rescaling of the {\tt dev} predicted probabilities that maximized the F1 score at
the 0.5 threshold, and applying this rescaling to the {\tt test} predicted probabilities.

@@

Though not plotted above, we also ran a model lesioning exercise-level features
including client, session type, format, and exercise duration. This model
achieved an AUROC of $.787$, far lower than any other lesion. This points to the
importance of how memory is assess on performance, reflecting insights
from psychology such as the difference between recall and recognition memory
\cite{yonelinas2002nature}.

@@

Finally, if the task had required broader generalization, psychologically inspired 
features might have performed more competitively. In the SLAM
task, there is a large amount of labeled training data for every user and for
most words. This allows simple ID-based features to work because the past
history of a user will likely influence their future performance. However, with
 ID-based features there is no way to generalize to newly-encountered users or 
 words, which have an ID that was not in the training set. The learned 
ID-based knowledge is useless here because there is no way to generalize from one 
unique ID to another. Theory-driven features, in
contrast, can often generalize to new settings because they capture aspects that 
are shared across (subsets of) users, words, or situations of the learning task.  For example, if we were asked to generalize to
a completely new language such as German, many parts of our model would falter
but word frequency, age of acquisition, and Levenshtein distance to
first-language translation would still likely prove to be features which
have high predictive utility.

In sum, we believe that the Duolingo SLAM dataset and challenge provide
interesting opportunities for cognitive science and psychology. 
Large-scale, predictive challenges like this one
might be used to identify features or variables that are important for learning.
Then, complementary laboratory-scale studies can be conducted which
establish the causal status of such features through controlled experimentation.
Conversely, insights from controlled experiments can be used to generate
new features that aid predictive models on naturalistic datasets~\cite{Griffiths:2015aa,Goldstone:2016aa}. This type of
two-way interaction could lead to long-run improvements in both scientific
explanation and real-world prediction.
